\begin{abstract}
Generative models applied to proteins are poised to revolutionize the \emph{in
silico} design of novel proteins satisfying various functional and shape
constraints. However, such models are notoriously hard to evaluate. The
\acrfull{mmd}, a statistic used in a kernel two-sample test, has emerged as a
highly versatile evaluation metric used to evaluate generative graph models.
This versatility is due to the fact that any kernel, and, therefore, any
appropriate underlying data representation, can be used. This makes it relevant
for proteins, because they can be represented as graphs (obtained with various
parameters), as point clouds, and as sequences of amino acids. In this thesis,
we aim to evaluate the applicability of different representations, descriptor
functions and kernel combinations for use in \acrshort{mmd} to design relevant
metrics to evaluate generative models operating in the protein domain. Through a
set of graph-based and point cloud-based perturbation experiments, we first
evaluate various configurations of graph descriptors traditionally used to
evaluate generative graph models. Second, we expand the use of \acrshort{mmd} to
encompass previously unused configurations, such as (i) graph kernels, (ii)
novel protein descriptor functions tailored to evaluate structural properties of
proteins such has the dihedral angles histogram formed by each pair of amino
acid, and the histogram of pairwise distances between amino acids, and (iii)
kernels operating on topological descriptors of proteins. We find that there are
multiple metrics that have a high correlation to various perturbations, and
recommend which to use based on the various \emph{desiderata} of suitable metrics:
expressivity, robustness and efficiency.

% TODO:  Complement with findings from conclusion
\end{abstract}
